Inferring Galactic Parameters from Chemical Abundances with Simulation-Based Inference

### 2025-02-25 src/tex/ms.tex
[showyourwork] overleaf sync

% \abstract{}{}{}{}{} 
% 5 {} token are mandatory
 \abstract
  % context heading (optional)
  % {} leave it empty if necessary  
   {Galactic chemical abundances provide crucial insights into fundamental galactic parameters, such as the high-mass slope of the initial mass function (IMF) and the normalization of Type Ia supernova (SN\,Ia) rates. Constraining these parameters is essential for advancing our understanding of stellar feedback, metal enrichment, and galaxy formation processes. However, traditional Bayesian inference techniques, such as Hamiltonian Monte Carlo (HMC), are computationally prohibitive when applied to large datasets of modern stellar surveys.}
  % aims heading (mandatory)
   {We leverage simulation-based-inference (SBI) as a scalable, robust, and efficient method for constraining galactic parameters from stellar chemical abundances and demonstrate its the advantages over HMC in terms of speed, scalability, and robustness against model misspecifications.}
  % methods heading (mandatory)
   {We combine a Galactic Chemical Evolution (GCE) model, \texttt{CHEMPY}, with a neural network emulator and a Neural Posterior Estimator (NPE) to train our SBI pipeline. Mock datasets are generated using \texttt{CHEMPY}, including scenarios with mismatched nucleosynthetic yields, with additional tests conducted on data from a simulated Milky Way-like galaxy. SBI results are benchmarked against HMC-based inference, focusing on computational performance, accuracy, and resilience to systematic discrepancies.}
  % results heading (mandatory)
   {SBI achieves a $\sim1,000\times$ speed-up compared to HMC, reducing inference runtime from $\gtrsim40$ hours to mere minutes for thousands of stars. Inference on $1,000$ stars yields precise estimates for the IMF slope ($\alpha_{\rm IMF} = -2.296 \pm 0.009$) and SN\,Ia normalization ($\log_{10}(N_{\rm Ia}) = -2.889 \pm 0.010$), deviating less than 0.04\% from the ground truth. SBI also demonstrates greater robustness to model misspecification than HMC, recovering accurate parameters even with alternate yield tables or data from a cosmological simulation.}
  % conclusions heading (optional), leave it empty if necessary 
   {SBI represents a paradigm shift in GCE studies, enabling efficient and precise analysis of massive stellar datasets. By outperforming HMC in speed, scalability, and robustness, SBI is poised to become a cornerstone methodology for future spectroscopic surveys facilitating deeper insights into the chemical and dynamical evolution of galaxies.}

@@

In essence, instead of computing the full model for each input parameter set, we pass the parameters to the NN which predicts the output abundances to high accuracy. As already argued in \cite{Philcox_2019} this has two benefits;
\begin{enumerate}
    \item \textbf{Speed:} The run-time of the \texttt{CHEMPY} function is $\sim1$\,s per input parameter set, which leads to very slow generation of training data for SBI. With the NN emulator, this reduces to $\sim5\times10^{-5}$\,s, and is trivially parallelizable, unlike \texttt{CHEMPY}.
    \item \textbf{Differentiability:} The NN is written in pytorch and has additionally a simple closed-form analytic structure \citep[described in the appendix of][]{Philcox_2019}, unlike the complex \texttt{CHEMPY} model. Both aspects allow it to be differentiated either via auto-diff or analytically, so one can use it to sample via advanced methods such HMC as done in \citep{Philcox_2019}.
\end{enumerate}

Despite the additional complexity introduced by using multiple stellar data-points, our NN simply needs to predict the birth-time abundances for a single star (with index $i$) from a given set of six parameters; $\{\vec\Lambda,\vec\Theta_i,T_i\}$. The same NN can be used for all $n_\mathrm{stars}$ stars (and run in parallel), reducing a set of $n_\mathrm{stars}$ runs of \texttt{CHEMPY} to a single matrix computation. With the above network parameter choices, the NN predicts abundances with an absolute percentage error of \variable{output/ape_NN.txt} (which translates into a logarithmic error of %\variable{output/ape_NN_log.txt} dex)  
which is far below typical observational errors and even smaller away from the extremes of parameter space (see Fig.~\ref{fig:ape_NN})\footnote{Our NN emulator is publicly available on the github repository accompanying this manuscript \href{https://github.com/TobiBu/sbi-chempy/blob/main/src/scripts/train_torch_chempy.py}{https://github.com/TobiBu/sbi-chempy/blob/main/src/scripts/train\_torch\_chempy.py} with pre-trained NN weights available on zenodo \href{https://zenodo.org/records/14507134}{https://zenodo.org/records/14507134}.}. In fact, we will add additional observational uncertainty to our mock observational data later during training of the neural posterior estimator network.

@@

In order to determine the optimal values of the global galactic parameters ($\Vec{\Lambda}$) one has to sample the posterior of Eq.~\ref{eq:posterior}. In practice this is a costly computation, since even with advanced techniques such as HMC sampling the posterior can only be evaluated for a small set of stars ($\lesssim200$) and requires long compute times \citep[$\sim40$ hours][]{Philcox_2019}.
%
However, recent advances in implicit-likelihood inference or SBI \citep{Cranmer2020} offer another very efficient approach to approximate the posterior (see next paragraph for more details). These methods train a neural conditional density estimator to represent the conditional posterior, $\mathbb{P}(\vec\Lambda|\{D_i\})$, which can be very efficiently evaluated given observational data $\{D_i\}$.

@@

We use a NPE \citep{zeghal2022neuralposteriorestimationdifferentiable} which utilises the gradients of the generative model $\mathcal{M}$ with a Neural Spline Flow \citep[NSF;][]{durkan2019neuralsplineflows} containing 20 hidden features and 10 transformation layers for the normalizing flow. The NSF uses as invertible transformation smooth and flexible spline functions to allow efficient sampling. The expressivity of the spline allows the NDE to be capture complex distributions, while also maintaining computational tractability. The final model was selected after extensive hyperparameter tuning, varying: the architecture between NSF, Masked Autoregressive Flow \citep[MAF;][]{papamakarios2018maskedautoregressiveflowdensity}, and Mask Autoregressive 
Flow with rational-quadratic spline (MAF-RQS), the number of neurons between {10, 20, 50, 100}, and the number of transformations between {1, 5, 10, 30}, optimizing over the test set for both the highest mean log posterior probability and the best calibration as measured by the TARP value \citep{Lemos2023}. For more details on the TARP value see section \ref{sec:sbc} in the appendix.

We train our NPE with $10^5$ data points consisting of $n_{\rm elements}=8$ simulated with the NN emulator described in the previous section. Inputs are sampled from a uniform prior over the original $5\sigma$-range of the Gaussian priors shown in Tab. \ref{tab:priors} to provide better coverage of the parameter space. Training takes $\sim35$ minutes on an Apple M1 Max.
We evaluate the accuracy of the NPE using $5\times10^4$ validation data points from the original \texttt{CHEMPY} simulator. In order to mimic observational noise, we add 5\% observational uncertainties to the abundances simulated with \texttt{CHEMPY} before feeding them to our NPE.
We have made sure that our NPE is well calibrated and posterior distributions are trustable. For more details on simulation-based calibration see appendix \ref{sec:sbc}.

@@

For each individual observation consisting of the abundance measurements of a single star we sample the posterior estimate for all six parameters $\{\vec\Lambda,\vec\Theta_i,T_i\}$ with 1,000 points. This takes around $0.3s$ for each star, making it extremely fast to infer the parameters of a large amount of stars. Our method takes around $50$ seconds to build a posterior for all six parameters for a dataset size of 1,000 stars (each time sampling the single star posterior with 1,000 points and fitting for the Gaussian parameters). Our combined runtime for the NPE training plus sampling is then $2160$ seconds on an Apple M1 Max for 1,000 stars hence our methods making it more than $340$ times faster than current HMC methods which take around $40$h for only $200$ stars \citep{Philcox_2019}. But since our approach is amortized and we do not need to retrain our NPE model each time we want to make an inference, we in fact achieve an inference speed-up of a factor of $\sim14,400$.
Hence, shorter computing times make it feasible to use orders of magnitudes more observations.

@@

However, currently our NPE network is not particularly good at estimating ages from abundances alone. On average our inference for ages results in an APE of $\sim27\%$ which is slightly larger than the observational noise of $20\%$ that we add during the mock up of our data.

@@

In Fig.~\ref{fig:CHEMPY_TNG_sbi} we show the joint posterior for $\alpha_{IMF}$ and $\log_{10}(N_{Ia})$ for our inference. The red star indicates the ground truth value, the black dot shows our posterior mean and white contours indicate $1-3\sigma$ levels.
%
Using a sample of 1,000 stars we infer \variable{output/CHEMPY_TNG_sbi.txt} which deviates less than $\sim0.04\%$ from the ground truth value.

@@

\begin{table}[]
\caption{Alternative nucleosynthetic yield tables used for model misspecification tests.}
     \centering
     \begin{tabular}{c|c}
       Type & Yield Table \\
        \hline
         SN\,Ia & \citet{2003NuPhA.718..139T} \\
         %SN\,II & \citet{2013ARA&A..51..457N} \\
         AGB & \citet{2016ApJ...825...26K}
     \end{tabular}
\label{tab:chempy_ALT_yields}
\end{table}

@@

This isn't up to date
\textcolor{blue}{
In Fig.~\ref{fig:CHEMPY_alt_N_star_analysis} we show our inference results (blue) for this setup for varying number of stars and compare them against HMC results (red) from \citet{Philcox_2019}. We see that the SBI results are less biased than the HMC results, especially for $\alpha_{\rm IMF}$. In fact with 100 stars HMC inferences are about \textcolor{red}{calculate value} $\sigma$ away from the ground truth value for $\alpha_{\rm IMF}$ and $\log_{10}(N_{\rm Ia})$, respectively. Again, the joint posterior is shown in Fig.~\ref{fig:CHEMPY_alt_sbi} and shows that while individually parameter inferences look good, jointly taken the SBI results are on the edge of being $3\sigma$ away from the ground truth.
}

\textcolor{blue}{
Nevertheless, the performance of our inference is still very good. We see that an increasing number of observations helps to decrease the models uncertainty just as before. However, we also note that our inference is now slightly biased as the observational data does not match the training data. Looking at Fig.~\ref{fig:CHEMPY_alt_sbi} we see that our inference is still consistent with the ground truth within the $3\sigma$ levels. Comparing this to the HMC results (red band in Fig.~\ref{fig:CHEMPY_alt_N_star_analysis}), we see that SBI is performing much better when the model is misspecified and leads to significantly less biased results. This is -- next to drastically reduced compute times -- another a key advantage of our SBI method compared to more standard approaches such as HMC.
}

@@

We show our inference results for the TNG data set in Fig.~\ref{fig:TNG_N_star_analysis} and Fig.~\ref{fig:CHEMPY_TNG_sim_sbi}, respectively.
Again, we find that inference becomes better with increasing number of stars. Despite the drastic difference in chemical enrichment model between training and testing data, our SBI pipeline is impressively capable of inferring the correct posterior values. SBI results for $\alpha_{\rm IMF}$ are almost perfect while $\log_{10}(N_{\rm Ia})$ is slightly biased low. This is exactly opposite to the results of HMC where inference for $\alpha_{\rm IMF}$ is biased and results for $\log_{10}(N_{\rm Ia})$ are more in agreement. Hence, our SBI inference is on par with the HMC results and given their extreme computational advantage they actually supersede HMC.

Looking at the joint posteriors in Figs.~\ref{fig:CHEMPY_TNG_sbi}, \ref{fig:CHEMPY_alt_sbi} and \ref{fig:CHEMPY_TNG_sim_sbi} we see that also in this case SBI recovers the true parameters with less than $3\sigma$ bias.

@@

The key assumption underpinning our methodology -- that individual stars are independently sampled from their respective stellar environments -- is foundational for tractability. However, this assumption warrants closer examination. In reality, stellar abundances are correlated due to shared star formation histories, cumulative enrichment, and dynamical interactions. Future efforts should explore methodologies capable of incorporating such correlations to further refine the accuracy of SBI, potentially through hierarchical modeling or graph-based methods.

An exciting avenue for future research involves expanding SBI to directly infer empirical nucleosynthetic yields, which remain a major uncertainty in GCE models. In the current framework, \texttt{CHEMPY} relies on tabulated yields from theoretical studies, which may not fully represent the complex processes driving stellar enrichment. Adapting SBI to simultaneously infer galactic parameters and refine empirical yield tables would require changes to the simulator. Specifically, \texttt{CHEMPY} would need to incorporate parameterized yield modifications as part of its input space, allowing for flexible adjustments to enrichment rates during inference. This would also necessitate larger training datasets and enhanced validation techniques to ensure convergence. Such an approach could provide a unified framework for calibrating galactic models directly against observational data.

@@

@@

Our results underscore the transformative advantages of SBI over traditional methods like Hamiltonian Monte Carlo (HMC), marking a paradigm shift in galactic parameter inference:
\begin{itemize}
    \item Orders-of-magnitude speed-up: SBI dramatically reduces computational requirements, achieving runtime improvements exceeding 1,000-fold compared to HMC. For instance, while HMC requires $\sim 40\,\mathrm{hours}$ to infer parameters from just 200 stars, SBI completes inference on thousands of stars in mere minutes. This efficiency makes SBI uniquely suited for analyzing the massive datasets expected from next-generation spectroscopic surveys.
    \item Scalability: SBI's amortized nature allows it to scale seamlessly with the size of the dataset. By training a neural posterior estimator once, the method can be applied repeatedly at virtually no additional computational cost. This scalability is essential for leveraging the millions of stars that future surveys like 4MOST \citep{4most} and WEAVE \citep{weave} will provide, enabling precise population-level inferences.
    \item Robustness to model misspecifications: Unlike HMC, which shows significant biases when faced with discrepancies between model assumptions and data, SBI demonstrates remarkable robustness. \textcolor{blue}{Even under conditions of mismatched yield tables (should be updated)} or data generated from hydrodynamical simulations, SBI provides accurate and reliable results. This robustness ensures SBI's applicability in real-world scenarios where exact model fidelity cannot be guaranteed.
\end{itemize}

@@

@@

\section{Additional inference results}

\begin{figure*}
    \centering
    \includegraphics[width=\textwidth]{figures/Nstar_comp_different_prior.pdf}
    \vspace{-.5cm}
    \caption{Same as Fig.~\ref{fig:CHEMPY_TNG_N_star_analysis} but for mock data a created with parameters for $\alpha_{IMF}$ and $\log_{10}(N_{Ia})$.}
    \label{fig:N_star_analysis_different_prior}
    \script{additional_inference.py}
\end{figure*}

>>>>

% \abstract{}{}{}{}{} 
% 5 {} token are mandatory
 \abstract
  % context heading (optional)
  % {} leave it empty if necessary  
   {Galactic chemical abundances provide crucial insights into fundamental galactic parameters, such as the high-mass slope of the initial mass function (IMF) and the normalization of Type Ia supernova (SN\,Ia) rates. Constraining these parameters is essential for advancing our understanding of stellar feedback, metal enrichment, and galaxy formation processes. However, traditional Bayesian inference techniques, such as Hamiltonian Monte Carlo (HMC), are computationally prohibitive when applied to large datasets of modern stellar surveys.}
  % aims heading (mandatory)
   {We leverage simulation-based-inference (SBI) as a scalable, robust, and efficient method for constraining galactic parameters from stellar chemical abundances and demonstrate its the advantages over HMC in terms of speed, scalability, and robustness against model misspecifications.}
  % methods heading (mandatory)
   {We combine a Galactic Chemical Evolution (GCE) model, \texttt{CHEMPY}, with a neural network emulator and a Neural Posterior Estimator (NPE) to train our SBI pipeline. Mock datasets are generated using \texttt{CHEMPY}, including scenarios with mismatched nucleosynthetic yields, with additional tests conducted on data from a simulated Milky Way-like galaxy. SBI results are benchmarked against HMC-based inference, focusing on computational performance, accuracy, and resilience to systematic discrepancies.}
  % results heading (mandatory)
   {SBI achieves a $\sim75,600\times$ speed-up compared to HMC, reducing inference runtime from $\gtrsim42$ hours to mere seconds for thousands of stars. Inference on $1,000$ stars yields precise estimates for the IMF slope ($\alpha_{\rm IMF} = -2.298 \pm 0.002$) and SN\,Ia normalization ($\log_{10}(N_{\rm Ia}) = -2.885 \pm 0.003$), deviating less than 0.05\% from the ground truth. SBI also demonstrates similar robustness to model misspecification than HMC, recovering accurate parameters even with alternate yield tables or data from a cosmological simulation.}
  % conclusions heading (optional), leave it empty if necessary 
   {SBI represents a paradigm shift in GCE studies, enabling efficient and precise analysis of massive stellar datasets. By outperforming HMC in speed, scalability, and robustness, SBI is poised to become a cornerstone methodology for future spectroscopic surveys facilitating deeper insights into the chemical and dynamical evolution of galaxies.}

@@

In essence, instead of computing the full model for each input parameter set, we pass the parameters to the NN which predicts the output abundances to high accuracy. As already argued in \cite{Philcox_2019} this has two benefits;
\begin{enumerate}
    \item \textbf{Speed:} The run-time of the \texttt{CHEMPY} function is $\sim1$\,s per input parameter set, which leads to very slow generation of training data for SBI. With the NN emulator, this reduces to $\sim5\times10^{-5}$\,s, and is trivially parallelizable, unlike \texttt{CHEMPY}. Having access to a fast simulator opens the possibility for testing Sequential Neural Posterior Estimate (SNPE) as an alternative, but we discuss this possibility in Section \ref{sec: discussion}. 
    \item \textbf{Differentiability:} The NN is written in pytorch and has additionally a simple closed-form analytic structure \citep[described in the appendix of][]{Philcox_2019}, unlike the complex \texttt{CHEMPY} model. Both aspects allow it to be differentiated either via auto-diff or analytically, so one can use it to sample via advanced methods such HMC as done in \citep{Philcox_2019}.
\end{enumerate}

Despite the additional complexity introduced by using multiple stellar data-points, our NN simply needs to predict the birth-time abundances for a single star (with index $i$) from a given set of six parameters; $\{\vec\Lambda,\vec\Theta_i,T_i\}$. The same NN can be used for all $n_\mathrm{stars}$ stars (and run in parallel), reducing a set of $n_\mathrm{stars}$ runs of \texttt{CHEMPY} to a single matrix computation. With the above network parameter choices, the NN predicts abundances with an absolute percentage error of \variable{output/ape_NN.txt} (which translates into a logarithmic error of %\variable{output/ape_NN_log.txt} dex)  
which is far below typical observational errors and even smaller away from the extremes of parameter space (see Fig.~\ref{fig:ape_NN})\footnote{Our NN emulator is publicly available on the github repository accompanying this manuscript \href{https://github.com/TobiBu/sbi-chempy/blob/main/src/scripts/train_torch_chempy.py}{https://github.com/TobiBu/sbi-chempy/blob/main/src/scripts/train\_torch\_chempy.py} with pre-trained NN weights available on zenodo \href{https://zenodo.org/records/14925307}{https://zenodo.org/records/14925307}.}. In fact, we will add additional observational uncertainty to our mock observational data later during training of the neural posterior estimator network.

@@

In order to determine the optimal values of the global galactic parameters ($\Vec{\Lambda}$) one has to sample the posterior of Eq.~\ref{eq:posterior}. In practice this is a costly computation, since even with advanced techniques such as HMC sampling the posterior can only be evaluated for a small set of stars ($\lesssim200$) and requires long compute times \citep[$\sim42$ hours][]{Philcox_2019}.
%
However, recent advances in implicit-likelihood inference or SBI \citep{Cranmer2020} offer another very efficient approach to approximate the posterior (see next paragraph for more details). These methods train a neural conditional density estimator to represent the conditional posterior, $\mathbb{P}(\vec\Lambda|\{D_i\})$, which can be very efficiently evaluated given observational data $\{D_i\}$.

@@

We use a NPE \citep{zeghal2022neuralposteriorestimationdifferentiable} which utilises the gradients of the generative model $\mathcal{M}$ with a Masked Autoregressive Flow 
\citep[MAF;][]{papamakarios2018maskedautoregressiveflowdensity} 
containing 8 hidden features and 4 transformation layers for the normalizing flow. 
%The NSF uses as invertible transformation smooth and flexible spline functions to allow efficient sampling. 
The expressivity of the MAF allows the NDE to be capture complex distributions, while also maintaining computational tractability. The final model was selected after extensive hyperparameter tuning, varying: the architecture between Neural Spline Flow \citep[NSF;][]{durkan2019neuralsplineflows}, MAF, and MAF with rational-quadratic spline (MAF-RQS), the number of neurons between {10, 20, 50, 100}, and the number of transformations between {1, 5, 10, 30}, optimizing over the test set for both the highest mean log posterior probability and the best calibration as measured by the TARP value \citep{Lemos2023}. For more details on the TARP value see section \ref{sec:sbc} in the appendix.

We train our NPE with $10^5$ data points consisting of $n_{\rm elements}=8$ simulated with the NN emulator described in the previous section. Inputs are sampled from the Gaussian priors shown in Tab. \ref{tab:priors}. Training takes $\sim10$ minutes on an Apple M1 Max.
We evaluate the accuracy of the NPE using $5\times10^3$ validation data points from the original \texttt{CHEMPY} simulator. In order to mimic observational noise, we add 5\% observational uncertainties to the abundances simulated with \texttt{CHEMPY} before feeding them to our NPE.
We have made sure that our NPE is well calibrated and posterior distributions are trustable. For more details on simulation-based calibration see appendix \ref{sec:sbc}.

@@

For each individual observation consisting of the abundance measurements of a single star we sample the posterior estimate for all six parameters $\{\vec\Lambda,\vec\Theta_i,T_i\}$ with 1,000 points. This takes around $0.3s$ for each star, making it extremely fast to infer the parameters of a large amount of stars. Our method takes around $10$ seconds to build a posterior for all six parameters for a dataset size of 1,000 stars (each time sampling the single star posterior with 1,000 points and fitting for the Gaussian parameters). Our combined runtime for the NPE training plus sampling is then $610$ seconds on an Apple M1 Max for 1,000 stars hence our methods making it more than $1240$ times faster than current HMC methods which take around $42$h for only $200$ stars \citep{Philcox_2019}. Note though, that since our approach is amortized and we do not need to retrain our NPE model each time we want to make an inference, we in fact achieve an inference speed-up of a factor of $\sim75,600$.
Hence, shorter computing times make it feasible to use orders of magnitudes more observations.

@@

However, currently our NPE network is not particularly good at estimating ages from abundances alone. On average our inference for ages results in an APE of $\sim27\%$ which is slightly larger than the observational noise of $20\%$ that we add during the mock up of our data but well in agreement with current uncertainties of stellar age inference.

@@

In Fig.~\ref{fig:CHEMPY_TNG_sbi} we show the joint posterior for $\alpha_{IMF}$ and $\log_{10}(N_{Ia})$ for our inference. The red star indicates the ground truth value, the black dot shows our posterior mean and white contours indicate $1-3\sigma$ levels.
%
Using a sample of 1,000 stars we infer \variable{output/CHEMPY_TNG_sbi.txt} which deviates less than $\sim0.04\%$ from the ground truth value. We have further checked the accuracy of our inference for a vastly different mock observational dataset created with shifted parameters of $\alpha_{IMF}=-2.1$ and $\log_{10}(N_{Ia})=-3$ and found that also in this case our model is well able to lead to correct inferences (see sec.~\ref{sec: additional inference} in the appendix).

@@

\begin{table}[]
\caption{Alternative nucleosynthetic yield tables used for model misspecification tests.}
     \centering
     \begin{tabular}{c|c}
       Type & Yield Table \\
        \hline
         SN\,Ia & \citet{2003NuPhA.718..139T} \\
         SN\,II & \citet{Nomoto2013} \\
         AGB & \citet{2016ApJ...825...26K}
     \end{tabular}
\label{tab:chempy_ALT_yields}
\end{table}

@@

In Fig.~\ref{fig:CHEMPY_alt_N_star_analysis} we show our inference results (blue) for this setup for varying number of stars and compare them against HMC results (red) from \citet{Philcox_2019}. We see that the SBI results are similar biased as the HMC results. In fact with 100 stars HMC inferences are about 4 and 3 $\sigma$ away from the ground truth value for $\alpha_{\rm IMF}$ and $\log_{10}(N_{\rm Ia})$, respectively. Again, the joint posterior is shown in Fig.~\ref{fig:CHEMPY_alt_sbi} and shows that while individually parameter inferences look good, jointly taken the SBI results are on the edge of being $3\sigma$ away from the ground truth.

Nevertheless, the performance of our inference is still very good. We see that an increasing number of observations helps to decrease the models uncertainty just as before. However, we also note that our inference is now slightly biased as the observational data does not match the training data. Looking at Fig.~\ref{fig:CHEMPY_alt_sbi} we see that our inference is inconsistent with the ground truth within several $\sigma$ levels. Comparing this to the HMC results (red band in Fig.~\ref{fig:CHEMPY_alt_N_star_analysis}), we see that SBI is performing similarly bad as HMC when the model is misspecified. Part of the problem here is that through our assumption of a factorized posterior, we decrease the inference uncertainty as we increase the number of observations. In future we will improve upon this through different model architectures. In summary, the drastically reduced compute times offer a key advantage of our SBI method compared to more standard approaches such as HMC. Nevertheless, in the limit of large star counts our current approach becomes over-confident and model uncertainities are underestimated.

@@

We show our inference results for the TNG data set in Fig.~\ref{fig:TNG_N_star_analysis} and Fig.~\ref{fig:CHEMPY_TNG_sim_sbi}, respectively.
Again, we find that inference becomes better with increasing number of stars. Despite the drastic difference in chemical enrichment model between training and testing data, our SBI pipeline is impressively capable of inferring the correct posterior values. SBI results for  $\log_{10}(N_{\rm Ia})$ are almost perfect while $\alpha_{\rm IMF}$ is slightly biased high. This is very similar to the results of HMC where inference for $\alpha_{\rm IMF}$ is biased and results for $\log_{10}(N_{\rm Ia})$ are more in agreement. Hence, our SBI inference is on par with the HMC results and given their extreme computational advantage they actually supersede HMC.

Looking at the joint posteriors in Figs.~\ref{fig:CHEMPY_TNG_sbi}, \ref{fig:CHEMPY_alt_sbi} and \ref{fig:CHEMPY_TNG_sim_sbi} we see that also in this case SBI recovers the true parameters only with a few $\sigma$ bias similar to HMC. Again, part of this problem is our assumpotion of a factorized posterior that underestimated model uncertainties as for the alternative yield case.

@@

The key assumption underpinning our methodology -- that individual stars are independently sampled from their respective stellar environments -- is foundational for tractability. However, this assumption warrants closer examination. In reality, stellar abundances are correlated due to shared star formation histories, cumulative enrichment, and dynamical interactions. Future efforts should explore methodologies capable of incorporating such correlations to further refine the accuracy of SBI, potentially through hierarchical modeling or graph-based methods. In particular this assumption leads to over-confident models in the limit of large observational samples that drastically underestimate model uncertainties. One very promising avenue to circumvent this probelm will be compositional score models \citep{compositional_score} which we will explore in future work.

An exciting avenue for future research involves expanding the SBI method presented here to be able to cope with missing data points by switching to a transformer model architecture.
Also, with slight modifications, SBI could be used 
to directly infer empirical nucleosynthetic yields, which remain a major uncertainty in GCE models. In the current framework, \texttt{CHEMPY} relies on tabulated yields from theoretical studies, which may not fully represent the complex processes driving stellar enrichment. Adapting SBI to simultaneously infer galactic parameters and refine empirical yield tables would require changes to the simulator. Specifically, \texttt{CHEMPY} would need to incorporate parameterized yield modifications as part of its input space, allowing for flexible adjustments to enrichment rates during inference. This would also necessitate larger training datasets and enhanced validation techniques to ensure convergence. Such an approach could provide a unified framework for calibrating galactic models directly against observational data. Research in this direction is currently performed and will be part of a future paper.

In the meantime, SBI can also be used to perform model comparison \citep[e.g.][]{model_comp,zhou2024} which can be used to determine which of the tabulated yield sets best match observational constraints.

@@

As anticipated in Section \ref{sec:methods}, the speed of our NN emulator would allow us to perform SNPE, giving up the amortized property in favor of a stronger constraining power, since the Sequential techniques have been shown empirically to outperform the respective amortized version \citep{Ho2024}. This method trains the NDE on a fraction of the initial simulations budget, retrieving a first estimate of the posterior, and at inference time simulate new observations to train on in regions of high posterior density, obtained from the first estimate. In this way we can obtain posteriors that are deliberately optimized for a singular data point. We have decided to leave this approach for future work because the results are quite promising and the amortized property can be crucial for the scalability.

@@

Our results underscore the transformative advantages of SBI over traditional methods like Hamiltonian Monte Carlo (HMC), marking a paradigm shift in galactic parameter inference:
\begin{itemize}
    \item Orders-of-magnitude speed-up: SBI dramatically reduces computational requirements, achieving runtime improvements exceeding 75,000-fold compared to HMC. For instance, while HMC requires $\sim 42\,\mathrm{hours}$ to infer parameters from just 200 stars, SBI completes inference on thousands of stars in mere minutes. This efficiency makes SBI uniquely suited for analyzing the massive datasets expected from next-generation spectroscopic surveys.
    \item Scalability: SBI's amortized nature allows it to scale seamlessly with the size of the dataset. By training a neural posterior estimator once, the method can be applied repeatedly at virtually no additional computational cost. This scalability is essential for leveraging the millions of stars that future surveys like 4MOST \citep{4most} and WEAVE \citep{weave} will provide, enabling precise population-level inferences.
    \item Robustness to model misspecifications: Unlike HMC, which shows significant biases when faced with discrepancies between model assumptions and data, SBI demonstrates remarkable robustness. Even under conditions of mismatched yield tables or data generated from hydrodynamical simulations, SBI provides accurate and reliable results. This robustness ensures SBI's applicability in real-world scenarios where exact model fidelity cannot be guaranteed.
\end{itemize}

@@

We note though, that the SBI pipeline presented here is not perfect and suffers from simplified assumptions that have been made in order to make the problem tracktable. In particular, the assumption that the  posterior factorizes is a strong assumption that leads to over-confidence of the models at large observational samples. This is certainly something that needs to be improved in future iterations and methodological work in this direction is currently undergoing.

@@

\section{Additional inference results}
\label{sec: additional inference}

\begin{figure*}
    \centering
    \includegraphics[width=\textwidth]{figures/Nstar_comp_different_prior.pdf}
    \vspace{-.5cm}
    \caption{Same as Fig.~\ref{fig:CHEMPY_TNG_N_star_analysis} but for mock data created with different parameters for $\alpha_{IMF}=-2.1$ and $\log_{10}(N_{Ia})=-3.0$.}
    \label{fig:N_star_analysis_different_prior}
    \script{additional_inference.py}
\end{figure*}

\begin{figure}
%\vspace{-1.3cm}
    \centering
    \includegraphics[width=\columnwidth]{figures/different_prior.png}
    \vspace{-.25cm}
    \caption{Same as Fig.~\ref{fig:CHEMPY_TNG_sim_sbi} but for the mock data taken created with a different parameter combination for $\alpha_{IMF}=-2.1$ and $\log_{10}(N_{Ia})=-3.0$.}
    \label{fig:CHEMPY_different_prior_sbi} 
    \script{additional_inference.py}
\end{figure}
